\documentclass[11pt,a4paper]{article}

% ── Packages ──────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=2.5cm]{geometry}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{array}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{tcolorbox}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{titlesec}

% ── Colors (Okabe–Ito colorblind-safe palette) ───────────────────────────
\definecolor{cOrange}{HTML}{E69F00}
\definecolor{cBlue}{HTML}{0072B2}
\definecolor{cSky}{HTML}{56B4E9}
\definecolor{cVermillion}{HTML}{D55E00}
\definecolor{cPurple}{HTML}{CC79A7}
\definecolor{cGray}{HTML}{999999}

% ── tcolorbox styles ─────────────────────────────────────────────────────
\tcbuselibrary{skins,breakable}

\newtcolorbox{keyinsight}[1][]{%
  enhanced, breakable,
  colback=cSky!8, colframe=cBlue,
  fonttitle=\bfseries, title=Key Insight,
  left=4pt, right=4pt, top=4pt, bottom=4pt, #1}

\newtcolorbox{managerbox}[1][]{%
  enhanced, breakable,
  colback=cOrange!8, colframe=cOrange,
  fonttitle=\bfseries, title=Manager's Briefing,
  left=4pt, right=4pt, top=4pt, bottom=4pt, #1}

\newtcolorbox{pitfall}[1][]{%
  enhanced, breakable,
  colback=cVermillion!8, colframe=cVermillion,
  fonttitle=\bfseries, title=Common Pitfall,
  left=4pt, right=4pt, top=4pt, bottom=4pt, #1}

% ── Header / Footer ──────────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\lhead{\small Stable Diffusion from Scratch}
\rhead{\small A Decision-Maker's Guide}
\cfoot{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ── Section styling ──────────────────────────────────────────────────────
\titleformat{\section}{\Large\bfseries\color{cBlue}}{\thesection}{1em}{}
\titleformat{\subsection}{\large\bfseries\color{cBlue!80!black}}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalsize\bfseries\color{cBlue!60!black}}{\thesubsubsection}{1em}{}

\hypersetup{
  colorlinks=true,
  linkcolor=cBlue,
  urlcolor=cBlue!80!black
}

% ══════════════════════════════════════════════════════════════════════════
\begin{document}

\begin{titlepage}
\centering
\vspace*{3cm}
{\Huge\bfseries Stable Diffusion from Scratch\\[0.4em]}
{\LARGE\color{cBlue} A Decision-Maker's Guide}

\vspace{2cm}
{\Large Lecture Notes}

\vspace{1.5cm}
{\large CAS BMAI --- ETH Z\"urich}

\vspace{3cm}
\begin{tabular}{rl}
\textbf{Format:} & 5 interactive notebooks, approx.\ 2.5 hours total \\
\textbf{Prerequisites:} & None (ability to click ``Run'' in Google Colab) \\
\textbf{Audience:} & Managers and non-technical leaders \\
\end{tabular}

\vfill
{\small \today}
\end{titlepage}

\tableofcontents
\newpage

% ══════════════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}

Text-to-image generation---the ability to produce photorealistic pictures
from a short textual description---has moved from a research curiosity to a
strategic capability in less than three years.  Products such as DALL-E,
Midjourney, and Stable Diffusion now power content pipelines in marketing,
e-commerce, entertainment, and product design.

These lecture notes accompany a hands-on course in which participants
\emph{build} a miniature text-to-image system themselves, step by step.
The system is small enough to train on a laptop (approximately 300\,K
parameters) yet architecturally identical to the full-scale models used
in production.  The goal is not to turn managers into machine-learning
engineers but to equip them with a \emph{working mental model} of the
technology---one precise enough to evaluate vendors, diagnose failures,
and ask engineering teams the right questions.

\subsection{The Co-Pilot Metaphor}

Throughout the course we use a single, consistent analogy that maps every
technical component to a familiar concept:

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{>{\bfseries}l l}
\toprule
Technical Component & Metaphor \\
\midrule
Diffusion model (U-Net) & The \emph{driver}---knows how to navigate but not where to go \\
Text prompt & The \emph{destination} \\
CLIP text encoder & The \emph{co-pilot} reading the map \\
Cross-attention & The driver asking the co-pilot at every turn \\
Guidance scale & How \emph{loudly} the co-pilot speaks \\
Negative prompt & The co-pilot saying ``NOT that way'' \\
Random seed & The random road network \\
\bottomrule
\end{tabular}
\end{center}

By the end of Notebook~5 you will have assembled all parts of this
``car'' and driven it yourself.

\subsection{Toy Dataset}

All experiments use a \textbf{programmatically generated} dataset of
coloured geometric shapes on a 32$\times$32 pixel canvas:

\begin{itemize}[nosep]
  \item \textbf{Shapes:} circle, square, triangle
  \item \textbf{Colours:} red, blue, green
  \item \textbf{Base dataset:} 9 classes (3 shapes $\times$ 3 colours)
  \item \textbf{Extended dataset (Notebook~5):} 36 classes (adding
        small/large and centred/offset)
\end{itemize}

Because every image is generated with known ground truth, we can measure
model quality objectively at every stage---an advantage rarely available
with real-world data.


% ══════════════════════════════════════════════════════════════════════════
\section{Notebook 1: The Raw Ingredients---Images and Text as Numbers}
\label{sec:nb1}

\textit{Approx.\ 25 minutes.  Metaphor: ``Here are the parts of the car
and the map.''}

\subsection{Cold Open: The End Product}

The notebook opens with a striking demonstration: a pre-trained model
generates coloured shapes from two-word prompts such as ``red circle'' or
``blue square.''  Participants see the finished product \emph{before}
learning how it works---establishing a concrete target for the sessions
ahead.

\subsection{What Computers See: Images as Tensors}

A digital image is stored as a three-dimensional grid of numbers---a
\textbf{tensor} of shape $H \times W \times C$, where $H$ is the height
in pixels, $W$ the width, and $C$ the number of colour channels (three
for RGB).  Each entry is a value between 0 and 1 representing the
intensity of one colour at one pixel location.

In the interactive exercise, participants adjust \textbf{brightness} and
\textbf{contrast} sliders and watch the raw numbers update in real time,
building intuition for how visual changes correspond to numerical changes.

\subsection{Words as Numbers: Embedding Vectors}

Computers cannot process words directly.  Instead, every word is mapped
to a list of numbers called an \textbf{embedding vector}.  Words with
similar meanings end up close together in this numerical space; unrelated
words end up far apart.

The notebook visualises nine word embeddings (colours and shapes)
projected onto a two-dimensional plane, letting participants see that
``red,'' ``blue,'' and ``green'' cluster together, while ``circle,''
``square,'' and ``triangle'' form a separate cluster.

The notion of ``closeness'' is quantified by \textbf{cosine similarity}:
two vectors pointing in roughly the same direction score near~1;
orthogonal vectors score~0.

\subsection{Productive Failure: The Random Projection}

Participants attempt to compare raw pixel vectors to text embeddings
using a na\"ive random projection.  The resulting similarity matrix shows
a random pattern with no meaningful structure.

\begin{pitfall}
Image pixels and word embeddings live in completely different numerical
spaces.  A random projection cannot bridge them---the similarity matrix
looks like noise.  Bridging this gap requires \emph{joint training}, as
we will see in Notebook~2.
\end{pitfall}

\begin{managerbox}
\textbf{Resolution = Cost.}  Our toy images are $32 \times 32 = 3{,}072$
numbers.  Stable Diffusion operates at $512 \times 512 = 786{,}432$
numbers; SDXL at $1024 \times 1024 = 3{,}145{,}728$ numbers.  Higher
resolution means proportionally more computation and cost.

\textbf{Vendor question:} ``What resolution does your model generate
natively?''
\end{managerbox}

\begin{managerbox}
\textbf{Joint vs.\ Post-Hoc Alignment.}  One could train an image system
and a text system separately, then bolt them together after the fact
(\emph{post-hoc} alignment).  As the random-projection failure
demonstrates, this does not work.  Effective systems train text and image
encoders \emph{jointly} from the start---the approach taken by CLIP.

\textbf{Vendor question:} ``Was your text encoder trained jointly with
your image model?''
\end{managerbox}


% ══════════════════════════════════════════════════════════════════════════
\section{Notebook 2: The Translator---CLIP}
\label{sec:nb2}

\textit{Approx.\ 30 minutes.  Metaphor: ``The co-pilot learns to read
the map.''}

\subsection{The Core Idea: Contrastive Learning}

\textbf{CLIP} (Contrastive Language--Image Pre-training) learns to place
images and their corresponding text descriptions \emph{near each other}
in a shared numerical space, while pushing mismatched pairs apart.

Training proceeds as follows:

\begin{enumerate}[nosep]
  \item Collect a batch of image--text pairs.
  \item Pass each image through an \textbf{image encoder} to obtain an
        image embedding.
  \item Pass each text through a \textbf{text encoder} to obtain a text
        embedding.
  \item Compute the cosine similarity between every image embedding and
        every text embedding in the batch.
  \item Adjust the encoders so that \emph{matching} pairs score high and
        \emph{non-matching} pairs score low.
\end{enumerate}

The loss function that drives this adjustment is called \textbf{InfoNCE}
(with a temperature parameter $\tau = 0.07$).  It is computed
symmetrically: text-to-image and image-to-text, averaged together.

\begin{keyinsight}
After training, CLIP provides a \emph{common language} for images and
text.  You can compare an image to a sentence---or two images, or two
sentences---simply by measuring the cosine similarity of their embedding
vectors.
\end{keyinsight}

\subsection{Watching CLIP Learn}

An interactive \textbf{epoch slider} (0 to 100) shows a PCA projection
of both text embeddings (orange circles) and image embeddings (blue
squares).  At epoch~0, matching pairs are scattered randomly.  As
training progresses, matching text and image embeddings migrate toward
each other until, by epoch~100, all nine pairs are tightly aligned.

\subsection{Building Mini-CLIP}

Participants inspect two small neural networks:

\begin{itemize}[nosep]
  \item \textbf{Text encoder:} word embedding (vocabulary of 12 tokens,
        32 dimensions) $\to$ mean pooling $\to$ two linear layers with
        ReLU $\to$ L2 normalisation.
  \item \textbf{Image encoder:} flatten $32 \times 32 \times 3 = 3{,}072$
        pixels $\to$ two linear layers with ReLU $\to$ L2 normalisation.
\end{itemize}

The complete Mini-CLIP model has approximately 200K parameters and trains
in $\sim$15 seconds on a CPU.

\subsection{Productive Failure: Memorisation vs.\ Generalisation}

When Mini-CLIP is trained on only \textbf{three} image--text pairs (red
circle, blue square, green triangle), it memorises them perfectly---the
similarity matrix shows three bright spots on the diagonal.  But it fails
completely on the other six classes: the off-diagonal entries are noise.

The model has \emph{memorised} individual pairs rather than learning
general concepts (``redness,'' ``circularity'').  Only when all nine pairs
are used does the model generalise.

\subsection{The Fix: Partial Matches Reveal Concept Learning}

With all nine training pairs, Mini-CLIP learns transferable concepts.
The text ``red square'' has moderate similarity to the image of a ``red
circle'' (shared colour), and ``blue circle'' scores well against ``green
circle'' (shared shape).  This \emph{partial-match} structure proves that
the model learned colour and shape as independent features---not just
rote associations.

\subsubsection*{Real-World Story: The CLIP Bias Problem}

Researchers discovered that CLIP embeddings encode societal biases present
in web-scraped training data.  Searching for ``CEO'' retrieves
predominantly white male faces; ``criminal'' retrieves disproportionately
Black men.  The diagnosis: contrastive learning faithfully encodes
correlations in the data---including unwanted ones.

\begin{managerbox}
\textbf{Fine-Tuning CLIP.}
\begin{itemize}[nosep]
  \item \textbf{Too few pairs} $\to$ memorisation.
  \item \textbf{Enough pairs} $\to$ concept learning.
  \item \textbf{Massive corpus} $\to$ rich, transferable knowledge.
\end{itemize}
\textbf{Vendor questions:} How large is the training set?  Has
demographic bias been tested?  How do you prevent catastrophic
forgetting when fine-tuning?
\end{managerbox}

\begin{managerbox}
\textbf{Practical Tool: Bias Auditing.}  You can audit a CLIP model by
computing cosine similarity between a concept embedding (e.g.,
``doctor,'' ``engineer,'' ``nurse'') and embeddings of demographically
varied images.  High variance in similarity scores across demographics
signals bias encoded in the model.
\end{managerbox}


% ══════════════════════════════════════════════════════════════════════════
\section{Notebook 3: The Engine---Diffusion}
\label{sec:nb3}

\textit{Approx.\ 35 minutes.  Metaphor: ``The driver learns to navigate
by being dropped at random locations.''}

\subsection{Forward Process: Adding Noise}

The diffusion framework rests on a surprisingly simple idea.  Start with
a clean image and gradually add random noise, step by step, until the
image becomes pure static.  This is the \textbf{forward process}---it is
easy, deterministic given the noise schedule, and requires no learning.

\subsection{Reverse Process: Denoising}

The creative act is running this process \emph{in reverse}: starting from
pure noise and removing noise step by step until a clean image emerges.
A neural network---the \textbf{denoiser}---is trained to predict the noise
that was added at each step.  Given a noisy image and the current step
number, it outputs an estimate of the noise, which is then subtracted.

Repeating this subtraction across all steps (50 in the demo) transforms
random static into a coherent image.

\subsection{The Cosine Noise Schedule}

The rate at which noise is added is controlled by a \textbf{noise
schedule}.  The course uses a \textbf{cosine schedule}, which adds noise
slowly at first, accelerates in the middle, and tapers at the end.  The
schedule is parameterised by cumulative noise levels $\bar{\alpha}_t$:
at $t = 0$, $\bar{\alpha}_0 = 1$ (clean image); at $t = T$,
$\bar{\alpha}_T \approx 0$ (pure noise).

An interactive \textbf{timestep slider} (0--50) lets participants watch a
single image progress from clean to noisy, with the schedule curve
displayed alongside.

\subsection{Noise Prediction Loss}

The training objective is straightforward: given a noisy image at
timestep~$t$, the denoiser predicts the noise that was added.  The loss
is the mean squared error between the predicted noise and the true noise.

\subsection{Productive Failure: The Na\"ive MLP}

Participants first train a \textbf{fully connected network (MLP)} as the
denoiser.  The MLP flattens the image into a single vector, destroying
all spatial structure.

The result: the MLP generates images indistinguishable from pure noise.
It has learned nothing useful because it cannot exploit the two-dimensional
structure of images.

\begin{pitfall}
A na\"ive MLP treats every pixel independently and has no notion of
spatial neighbourhood.  Flattening a 2D image into a 1D vector discards
exactly the structure needed for coherent image generation.
\end{pitfall}

\begin{keyinsight}
The architecture of the denoiser determines \emph{what kinds of
structure} the model can learn.  Choosing the right architecture is not a
minor implementation detail---it is a strategic decision that shapes
output quality.  The fix: switch to a \textbf{convolutional}
architecture (the U-Net, introduced in Notebook~4).
\end{keyinsight}

\subsubsection*{Real-World Story: The Deepfake Dilemma}

In 2023, AI-generated images were submitted as legal evidence.  They
looked photorealistic because photorealism is the explicit \emph{training
objective}---the model minimises the difference between generated and
real image distributions.  Detecting AI-generated content requires
specialised tools that exploit subtle frequency-domain artefacts left by
the generation process.

\begin{managerbox}
\textbf{Denoising Steps = Cost.}  Each denoising step requires a full
forward pass through the neural network.  Fifty steps means fifty times
the compute of a single image classification.

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ll}
\toprule
Setting & Steps \\
\midrule
Standard quality & 50 \\
Production (balanced) & 20--50 \\
Fast / distilled models & 1--4 \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Vendor questions:} How many denoising steps does your model use?
Is guidance computed in parallel or serially?  Can users trade quality
for speed?
\end{managerbox}


% ══════════════════════════════════════════════════════════════════════════
\section{Notebook 4: The Conversation---Cross-Attention}
\label{sec:nb4}

\textit{Approx.\ 30 minutes.  Metaphor: ``At every turn, the driver asks
the co-pilot.''}

\subsection{From Unconditional to Conditional Generation}

The diffusion model from Notebook~3 generates images, but it has no way
to steer \emph{what} it generates.  The driver can navigate but does not
know the destination.  We need a mechanism for the text prompt (via CLIP
embeddings) to influence the denoising process at every step.

\subsection{The MicroUNet Architecture}

The notebook introduces a small convolutional denoiser---the
\textbf{MicroUNet} ($\sim$160K parameters)---with the following
structure:

\begin{enumerate}[nosep]
  \item \textbf{Encoder:} three convolutional layers with GroupNorm,
        progressively downsampling the $32 \times 32$ image to an
        $8 \times 8$ feature map (128 channels).
  \item \textbf{Bottleneck:} timestep injection via a learned embedding,
        plus a \textbf{cross-attention layer} connecting image features to
        text features.
  \item \textbf{Decoder:} two transposed convolutions with skip
        connections from the encoder, restoring the original resolution.
\end{enumerate}

\subsection{How Cross-Attention Works}

Cross-attention is the mechanism by which image features ``consult'' text
features during denoising.  At the bottleneck of the U-Net:

\begin{enumerate}[nosep]
  \item The 64 image patches (8$\times$8 spatial grid, 128-dimensional
        features) produce \textbf{queries}~(Q): ``What should I look
        like here?''
  \item The per-token text embeddings (32-dimensional) produce
        \textbf{keys}~(K) and \textbf{values}~(V): ``Here is what the
        prompt says.''
  \item Each query is compared to all keys via a scaled dot-product
        similarity.
  \item The resulting \textbf{attention weights} determine how much each
        text token influences each spatial location.
  \item The weighted values are projected back to 128 dimensions and
        added as a \textbf{residual connection} to the image features.
\end{enumerate}

This process repeats at every denoising step, providing continuous
text guidance.

\begin{keyinsight}
Cross-attention is not a one-time instruction; it is a \emph{continuous
conversation} between the image being generated and the text prompt.  At
every denoising step and at every spatial location, the image ``asks''
the text what to do.
\end{keyinsight}

\subsection{Demo: Cross-Attention ON vs.\ OFF}

A side-by-side comparison shows the same model generating with
cross-attention enabled (sharp, text-aligned shapes) versus disabled
(blurry, generic output).  The contrast is immediate and dramatic.

\subsection{Visualising Attention Heatmaps}

Attention weights are visualised as heatmaps over the $8 \times 8$
spatial grid, one heatmap per text token.  Participants observe that the
token ``red'' activates the spatial locations where the shape appears,
while ``circle'' activates the shape's boundary region.

\subsection{Interactive Explorer: Guidance Scale}

A slider varies the \textbf{guidance scale} from 1.0 to 15.0, showing
five side-by-side generations:

\begin{itemize}[nosep]
  \item \textbf{Low scale ($\sim$1):} the co-pilot whispers---creative
        but unfocused output.
  \item \textbf{Sweet spot (7--10):} the co-pilot speaks clearly---sharp,
        text-faithful results.
  \item \textbf{High scale ($>$10):} the co-pilot shouts---follows
        instructions obsessively but may oversaturate.
\end{itemize}

\subsection{Productive Failure: Uniform Attention}

When attention weights are forced to be \emph{uniform} (every text token
influences every pixel equally---``co-pilot with earplugs''), the output
collapses into a blurry, generic blob that ignores the prompt entirely.
Restoring the learned attention weights immediately recovers sharp,
text-aligned generation.

\subsubsection*{Real-World Story: The E-Commerce Material Swap}

A company prompted a model with ``silver watch on wooden table.''  The
output: a wooden-looking watch on a metallic table.  This
\emph{attribute binding} failure arises because cross-attention has no
built-in constraint forcing attributes (``silver,'' ``wooden'') to bind
to the correct objects.

\begin{managerbox}
\textbf{Compositional Accuracy.}  Cross-attention enables text guidance
but suffers from the \emph{attribute binding problem}: when multiple
objects and attributes are present, the model can swap which attribute
belongs to which object.  There is no hard architectural constraint
preventing this.

\textbf{Vendor question:} ``Can you show me attention maps for a failed
generation?''  Attention visualisation reveals \emph{why} a model
produced a wrong output---invaluable for debugging.
\end{managerbox}


% ══════════════════════════════════════════════════════════════════════════
\section{Notebook 5: The Road Trip---The Full Pipeline}
\label{sec:nb5}

\textit{Approx.\ 30 minutes.  Metaphor: ``Start the engine, set the
destination, drive.''}

\subsection{Assembling the Pipeline}

In this final notebook, all components are wired together into a complete
text-to-image system:

\begin{enumerate}[nosep]
  \item The \textbf{text prompt} is encoded by the trained Mini-CLIP
        text encoder into per-token embeddings.
  \item A \textbf{random noise image} is sampled (determined by the
        seed).
  \item Over 50 denoising steps, the \textbf{MicroUNet} progressively
        removes noise, consulting the text embedding via
        \textbf{cross-attention} at each step.
  \item The final output is a clean, generated image matching the prompt.
\end{enumerate}

The climactic moment of the course: participants type a prompt such as
``red circle'' and watch the system they built transform pure noise into
a recognisable red circle, displayed as a filmstrip of intermediate
denoising frames.

\subsection{Classifier-Free Guidance}

At each denoising step, the model runs \textbf{two forward passes}:

\begin{enumerate}[nosep]
  \item \textbf{Conditional:} the U-Net receives the actual text
        embedding.
  \item \textbf{Unconditional:} the U-Net receives a null (empty) text
        embedding.
\end{enumerate}

The two predictions are blended:
\[
  \hat{\epsilon} = \epsilon_{\text{uncond}} + s \cdot
  (\epsilon_{\text{cond}} - \epsilon_{\text{uncond}})
\]
where $s$ is the guidance scale.  A scale of $s = 1$ yields purely
unconditional generation; values of $s = 7$--$10$ produce sharp,
text-faithful images; very high values ($s > 15$) cause oversaturation
and artefacts.

In the co-pilot metaphor: the guidance scale controls how loudly the
co-pilot speaks.  Too quiet and the driver wanders; too loud and the
driver oversteers.

\subsection{Interactive Prompt Lab}

Participants experiment freely with:
\begin{itemize}[nosep]
  \item A \textbf{prompt dropdown} (all 9 base classes plus variations)
  \item A \textbf{seed slider} (0--999)
  \item A \textbf{guidance scale slider} (1.0--15.0)
\end{itemize}

Each generation displays a filmstrip of eight intermediate denoising
frames, making the progressive refinement process visible.

\subsection{Productive Failure: Compositional Complexity}

The dataset is expanded from 9 classes to \textbf{36 classes} (adding
two sizes and two positions).  The model's colour and shape accuracy
drops, especially when multiple attributes must bind correctly (e.g.,
``small red circle'' sometimes produces a small circle with red appearing
elsewhere).

This demonstrates the fundamental \textbf{combinatorial challenge}:
the number of possible attribute combinations grows multiplicatively,
but training data is finite.  Real-world models face the same scaling
problem at vastly larger scale.

\subsection{Safety: Where Would Filters Go?}

The notebook discusses where safety filters belong in the pipeline:

\begin{enumerate}[nosep]
  \item \textbf{Text input:} reject or rewrite blocked prompts before
        encoding.
  \item \textbf{CLIP embedding:} check the embedding vector against
        known harmful regions.
  \item \textbf{Final image:} classify the output and block unsafe
        content.
\end{enumerate}

Because text guidance operates at \emph{every} denoising step, a harmful
prompt influences the entire generation process---making input-level
filtering the most effective first line of defence.

\subsection{Bridge to Real Tools}

The toy model ($\sim$300K parameters) and production Stable Diffusion
($\sim$860M parameters, 512$\times$512 output) share the \emph{same
architecture}: CLIP text encoder, U-Net denoiser with cross-attention,
cosine noise schedule, and classifier-free guidance.  The only difference
is scale.  SDXL pushes further to $\sim$3.5B parameters and
$1024 \times 1024$ resolution.  Understanding the toy system transfers
directly to reasoning about production systems.

\subsubsection*{Real-World Story: The Prompt Engineering Economy}

Small wording changes---``a red circle'' versus ``circle, red''---produce
different CLIP embeddings and therefore different outputs.  A marketing
team discovered that switching from casual language to precise
descriptors (``product photography of blue running sneaker, centred,
white studio background\ldots'') dramatically improved consistency and
quality.  Prompt sensitivity is an \emph{architectural property}, not a
user error.

\begin{managerbox}
\textbf{Final Briefing.}  You have now built every component of a
text-to-image system at miniature scale.  Real Stable Diffusion is
roughly 25\,000$\times$ larger but \textbf{architecturally identical}.
When your engineering team says ``increase the guidance scale,'' you know
exactly what that means: two forward passes per step, blend the
predictions, amplify the text signal.  You understand the system
\emph{mechanically}, not vaguely.
\end{managerbox}


% ══════════════════════════════════════════════════════════════════════════
\section{The Co-Pilot Reference Card}
\label{sec:reference}

Each notebook adds one row to a cumulative reference table.  By the end
of Notebook~5, participants have a complete cheat sheet:

\begin{center}
\renewcommand{\arraystretch}{1.35}
\small
\begin{tabular}{@{} >{\bfseries}p{0.8cm} p{2.8cm} p{5cm} p{3.5cm} @{}}
\toprule
NB & Component & What It Does & Co-Pilot Metaphor \\
\midrule
1 & Raw Ingredients &
  Images = tensors of numbers; text = embedding vectors;
  cosine similarity measures closeness &
  ``Here are the parts of the car and the map'' \\
2 & CLIP (Translator) &
  Learns a shared embedding space; contrastive loss pulls
  matching pairs together &
  ``The co-pilot learns to read the map'' \\
3 & Diffusion (Engine) &
  Forward: add noise.\ Reverse: predict \& remove noise.
  U-Net preserves spatial structure &
  ``The driver learns to navigate by being dropped at random
  locations'' \\
4 & Cross-Attention (Conversation) &
  Q = image, K/V = text.\ Each image region checks the text.
  Guidance scale = co-pilot volume &
  ``At every turn, the driver asks the co-pilot'' \\
5 & Full Pipeline (Road Trip) &
  Wire it all together: noise $\to$ denoise with text
  guidance $\to$ image.\ CFG = two passes &
  ``Start the engine, set the destination, drive'' \\
\bottomrule
\end{tabular}
\end{center}


% ══════════════════════════════════════════════════════════════════════════
\section{Five Questions to Ask an AI Image Generation Vendor}
\label{sec:questions}

\begin{enumerate}
  \item \textbf{Resolution \& cost} (from NB\,1).  What resolution does
        your model generate natively, and what is the per-image compute
        cost?
  \item \textbf{Training data \& bias} (from NB\,2).  How large and
        diverse is your training dataset?  What bias audits have been
        conducted?
  \item \textbf{Architecture \& speed} (from NB\,3).  How many denoising
        steps does your model use?  Can users trade quality for speed?
  \item \textbf{Diagnostics} (from NB\,4).  Can you provide attention
        maps or other interpretability tools to diagnose failed
        generations?
  \item \textbf{Reproducibility} (from NB\,5).  Does your system log
        random seeds and model versions for every generated asset,
        enabling deterministic replay?
\end{enumerate}


% ══════════════════════════════════════════════════════════════════════════
\section{Vendor Evaluation: Decision Tree}
\label{sec:decision}

When adopting AI image generation, organisations face three broad
options:

\begin{description}[style=nextline, leftmargin=1.5cm]
  \item[API Service] Use a third-party API (e.g., OpenAI, Stability AI).
    Best when: speed to market is paramount, volume is moderate, and
    customisation needs are low.  Trade-off: the vendor controls model
    updates and data handling.
  \item[Fine-Tuned Model] Take a pre-trained open model and fine-tune it
    on proprietary data.  Best when: domain-specific quality matters
    (e.g., your brand's visual style) and you have in-house ML capacity.
    Trade-off: faster convergence but requires curating training data.
  \item[Self-Hosted Model] Deploy a model entirely on your own
    infrastructure.  Best when: data privacy is non-negotiable, you need
    full control, and you have significant engineering resources.
    Trade-off: you bear all compute cost and model maintenance.
\end{description}

The right choice depends on your organisation's constraints along four
axes: \textbf{data sensitivity}, \textbf{customisation depth},
\textbf{engineering capacity}, and \textbf{cost structure}.


% ══════════════════════════════════════════════════════════════════════════
\section{Jargon Decoder}
\label{sec:jargon}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\small
\begin{tabular}{@{} >{\bfseries}p{3.5cm} p{9cm} @{}}
\toprule
Term & Plain-Language Definition \\
\midrule
Tensor & A multi-dimensional grid of numbers (an image is a 3-D tensor) \\
Embedding & A list of numbers representing a word, sentence, or image in a form computers can process \\
Cosine similarity & A score measuring how similar two embeddings are (1 = identical direction, 0 = unrelated) \\
CLIP & A model that learns to place matching images and texts near each other in a shared number space \\
Contrastive learning & A training strategy that pushes matches together and mismatches apart \\
Diffusion & A generative process that creates images by iteratively removing noise from random static \\
U-Net & A convolutional neural network with skip connections, well suited for image denoising \\
Cross-attention & A mechanism allowing one type of data (images) to query another (text) for guidance \\
Guidance scale & A parameter controlling how strongly the text prompt influences image generation \\
Noise schedule & The plan governing how quickly noise is added or removed at each step \\
Latent space & A compressed numerical representation where images are manipulated more efficiently \\
VAE & Variational Auto-Encoder: compresses images into latent space and reconstructs them back \\
LoRA & Low-Rank Adaptation: a lightweight fine-tuning technique that modifies only a small fraction of model weights \\
Fine-tuning & Further training of an existing model on specialised data to improve domain-specific quality \\
Inpainting & Selectively regenerating a chosen region of an image while keeping the rest unchanged \\
\bottomrule
\end{tabular}
\end{center}


% ══════════════════════════════════════════════════════════════════════════
\section{Red Flags When Evaluating Vendors}
\label{sec:redflags}

\begin{enumerate}
  \item \textbf{``We use a proprietary architecture.''}  If the vendor
        cannot describe their model's architecture, you cannot evaluate
        its limitations or plan for adaptation.
  \item \textbf{No guidance scale parameter.}  Without user-controllable
        guidance, you cannot tune the balance between creativity and
        prompt fidelity for your use case.
  \item \textbf{Fixed single-step generation.}  A system that does not
        expose the number of denoising steps removes a key quality--speed
        trade-off lever.
  \item \textbf{No API for custom training.}  If the vendor offers no
        path to fine-tuning, you are permanently locked into their base
        model's capabilities and biases.
  \item \textbf{No bias or safety audit.}  The vendor cannot describe
        what biases exist in their training data, what safety filters are
        in place, or how content provenance is tracked.
\end{enumerate}


% ══════════════════════════════════════════════════════════════════════════
\section{Conclusion}
\label{sec:conclusion}

The core architecture of text-to-image generation rests on three
learnable components---a text--image alignment model (CLIP), a
convolutional denoising network (MicroUNet with diffusion), and a
conditioning mechanism (cross-attention)---plus a set of carefully chosen
hyper-parameters (cosine noise schedule, guidance scale, seed).

None of these pieces is, on its own, mysterious.  Images are grids of
numbers.  Text is lists of numbers.  CLIP aligns the two.  Diffusion
learns to remove noise.  Cross-attention lets text steer the denoiser.
Classifier-free guidance amplifies the text signal.  The compound system
produces results that feel like magic, but every step is auditable,
tuneable, and---as you have now experienced first-hand---buildable from
scratch.

Understanding these building blocks transforms the conversation with
engineering teams and vendors from ``Can the AI do this?''\ to ``Which
component is responsible for this limitation, and what would it take to
fix it?''

\vfill

\begin{center}
\small\color{cGray}
\textit{These lecture notes accompany the interactive course
``Stable Diffusion from Scratch: A Decision-Maker's Guide.''}\\
\textit{All interactive elements, code exercises, and pre-/post-course
quizzes are available in the companion Jupyter notebooks.}
\end{center}

\end{document}
